---
title: "Matthew Russell Code Sample"
output: html_document
---

# set up

```{r}
library(tidyverse)
library(poLCA)
library(gtsummary)
library(knitr)
library(lmtest)
library(ggthemes)
library(tidyLPA)
```

# read in data and filter so it's only pregnant observations

```{r}

# read in data

teds <- readr::read_csv("./tedsd_puf_2019.csv")

# filter data so it's only pregnant women

preg_df <-
  teds %>% 
  janitor::clean_names() %>%
  dplyr::mutate(
    id = seq(1:nrow(teds))
  ) %>% 
  dplyr::filter(preg == 1) %>% 
  dplyr::relocate(id, .before = disyr)

# remove teds data to make space

rm(teds)
gc()

```

# select relevant columns and convert all to meaningful factors

```{r}

# grab the ids from the
# among data only are looking  pregnant women with the variables we're interested 
# in using to build the latent classes
ids <- 
preg_df %>% 
  dplyr::select(id, alcflg, frstuse1, herflg, marflg, mthamflg, noprior, psyprob, reason) %>% 
  dplyr::filter(reason %in% c(1,2) & frstuse1 != -9 & noprior != -9 & psyprob != -9) %>% 
  na.omit() %>% 
  dplyr::pull(., id)

# create copy of preg_df
# where we filter to only the ids in ids vector
# convert variables to factors

preg_factor <-
  preg_df %>% 
  dplyr::filter(id %in% ids) %>% 
  dplyr::mutate(
    alcflg = factor(alcflg, levels = c(0,1), labels = c("No", "Yes")), 
    comp_trt = factor(ifelse(reason == 1, "Yes", "No")),
    frstuse1 = factor(frstuse1) %>% 
      forcats::fct_collapse(., 
                   "14 and under" = c("1", "2"),
                   "15 to 17" = "3", 
                   "18 to 20" = "4", 
                   "21+" = c("5", "6", "7")), 
    herflg = factor(herflg, levels = c(0,1), labels = c("No", "Yes")), 
    marflg = factor(marflg, levels = c(0,1), labels = c("No", "Yes")), 
    mthamflg = factor(mthamflg, levels = c(0,1), labels = c("No", "Yes")),
    prior = factor(noprior, levels = c(0:1), labels = c("No Treatment", "One or more treatment")), 
    psyprob = factor(psyprob, levels = c(1:2), labels = c("Yes", "No")) %>% forcats::fct_rev(.)
  ) %>% 
  dplyr::select(!c(reason, noprior)) %>% 
  dplyr::select(order(colnames(.))) %>% 
  dplyr::relocate(id, .before = alcflg) %>% 
  as.data.frame() %>% 
  na.omit()

# pull out treatment vector

comp_trt <- preg_factor %>% dplyr::pull(., comp_trt)

# subset to only independent variables

no_outcome_df <- preg_factor %>% 
  dplyr::select(alcflg, frstuse1, herflg, marflg, mthamflg, prior, psyprob)

```

# lca

```{r}

# create the formula of the variables to put into the LCA

attach(no_outcome_df)
formula <- cbind(alcflg, frstuse1, herflg,
         marflg, mthamflg, prior, psyprob) ~ 1
detach(no_outcome_df)

# run the LCA 
# from 1 class to 5 class models
# then compare AIC, BIC, and G-squared

lc_list <- purrr::map(1:5, function(x) {
  
  set.seed(100)
  poLCA::poLCA(formula, no_outcome_df, nclass = x, maxiter = 10000)
}
)

# compare aics, bics, and G2

tibble::tibble(
  num_class = seq(1:5), 
  aic = purrr::map_dbl(lc_list, "aic"),
  bic =  purrr::map_dbl(lc_list, "bic"), 
  `g-squared` = purrr::map_dbl(lc_list, "Gsq")
) %>% 
  dplyr::arrange(num_class) %>% 
  knitr::kable(digits = 2)

# get frequency information among our independent variables for our formula
no_outcome_df %>% 
  dplyr::select(c(alcflg, frstuse1, herflg,
         marflg, mthamflg, prior, psyprob)) %>% 
  gtsummary::tbl_summary(.)

```

*we're using a four-class model*

# adding predicted classes

```{r, adding predicted class}

pred_class_df <-
  preg_factor %>% 
  dplyr::filter(id %in% ids) %>% 
  dplyr::select(c(age, race, ethnic, marstat, educ, employ, livarag)) %>% 
  dplyr::mutate(
    pc_4 = factor(lc_list[[4]]$predclass), 
    comp_trt = comp_trt, 
    age = factor(age, levels = c(1:12), 
                 labels = c("12-14", "15-17", 
                            "18-20", "21-24", 
                            "25-29", "30-34", 
                            "35-39", "40-44", 
                            "45-49", "50-54", 
                            "55-64", "65+")) %>% 
      forcats::fct_collapse(., 
                   `12-29` = c("12-14", "15-17", 
                            "18-20", "21-24", 
                            "25-29"), 
                   `30+` = c("30-34", 
                            "35-39", "40-44", 
                            "45-49", "50-54", 
                            "55-64", "65+")),
    race = factor(race, levels = c(1:9)) %>% 
      forcats::fct_collapse(., 
                   "white" = "5", 
                   other_level = "other"), 
    ethnic = factor(ethnic, levels = c(1:5)) %>% 
      forcats::fct_collapse(.,
                   "not_hispanic" = "4", 
                   other_level = "hispanic"), 
    marstat = factor(marstat, levels = c(1:4),
                     labels = c("never_married", "married",
                                "separated", "divorced/widowed")), 
    educ = factor(educ, levels = c(1:5)) %>% 
      forcats::fct_collapse(., 
                   "less_than_hs" = c("1", "2"), 
                   "hs" = "3", 
                   other_level = "more_than_hs"), 
    employ = factor(employ, levels = c(1:4)) %>% 
      forcats::fct_collapse(., 
                   "employed" = c("1", "2"), 
                   "unemployed" = "3", 
                   other_level = "not_in_labor_force") %>% 
      forcats::fct_relevel(., "not_in_labor_force", "unemployed", "employed"), 
    livarag = factor(livarag, levels = c(1:3)) %>% 
      forcats::fct_collapse(., 
                   "not_homeless" = c("2", "3"),
                   other_level = "homeless")
  )

# print summary of demographic variables
pred_class_df %>% 
  dplyr::select(-pc_4) %>% 
  gtsummary::tbl_summary(.)


```

# logistic regression: four-class model

```{r}

# four class model 

pred_class_df$pc_4 <- relevel(pred_class_df$pc_4, ref = "4")

# null model

null_log_m_4 <- glm(comp_trt ~ pc_4, data = pred_class_df, family = "binomial")

# summary of null model
summary(null_log_m_4)

# coefficients and CI of null model 
exp(cbind("Odds ratio" = coef(null_log_m_4), confint.default(null_log_m_4, level = 0.95))) %>% 
  kable(digits = 4)

# full model

full_log_m_4 <- glm(comp_trt ~ age  + race + ethnic + marstat + educ + employ + livarag + pc_4, data = pred_class_df, family = "binomial")

# summary of full model
summary(full_log_m_4)

# coefficients and CI of full model 
exp(cbind("Odds ratio" = coef(full_log_m_4), confint.default(full_log_m_4, level = 0.95))) %>%
  kable(digits = 2)

# LRT for full vs null 

log_lrt <- glm(comp_trt ~ age  + race + ethnic + marstat + educ + employ + livarag, data = pred_class_df, family = "binomial")

lrtest(full_log_m_4, log_lrt)

```

# distribution of categorical variables by class: 4 class model

```{r}

# print demographic table summary for each of the classes
for (i in c(1:4)) {
  
  pred_class_df %>% 
  dplyr::select(-comp_trt) %>% 
  dplyr::filter(pc_4 == as.character(i)) %>% 
  gtsummary::tbl_summary(.) %>% print()

}

```

# frequencies for each level of the variable per class: 4 class model

```{r}
# this is data manipulation so that we can each row represents a 
# variable and a level and we can see the N sizes of each variable/level
# combination per class 

freq_per_var_level <-
  pred_class_df %>% 
    tidyr::pivot_longer(
      cols = !pc_4, 
      names_to = "variable", 
      values_to = "level"
    ) %>% 
    dplyr::rename(class = pc_4) %>% 
    dplyr::mutate(
      class = as.numeric(class)
    ) %>% 
    dplyr::select(variable, level, class) %>% 
    dplyr::group_by(variable, level, class) %>% 
    dplyr::mutate(
      freq_total = n()
    ) %>% 
    dplyr::ungroup() %>% 
    dplyr::group_by(class, variable) %>% 
    dplyr::mutate(
      class_total = n(), 
    ) %>% 
    dplyr::ungroup() %>% 
    dplyr::select(variable, level, class, freq_total) %>% 
    dplyr::distinct() %>%
    tidyr::pivot_wider(
      names_from = class, 
      values_from = freq_total, 
      names_prefix = "class_freq_"
    ) %>% 
    dplyr::select(variable, level, class_freq_1, class_freq_2, class_freq_3, class_freq_4) %>% 
    dplyr::arrange(variable, level) %>% 
  na.omit()

```

# prop test

```{r, prop test}

# grab the N sizes of each of the classes 
n_sizes <- 
pred_class_df %>% 
  dplyr::count(pc_4) %>% 
  dplyr::mutate(pc_4 = as.numeric(pc_4)) %>% 
  dplyr::arrange(pc_4) %>% 
  dplyr::pull(., n)

# initialize list where we will add p-values from prop.test
output <- list()

for (i in 1:20) {
  
  # run prop.test on each row of freq_per_var_level
  # using final four columns of freq_per_var_level as counts 
  # and using n_sizes as our n's
  output[[i]] = prop.test(c(t(freq_per_var_level[i, 3:6])), n_sizes)$p.value
  
}

# add vector of p-values for homogeneity of proportions to each test 

hom_prop_df <-
  freq_per_var_level %>% 
  dplyr::mutate(
    p_value = unlist(output)
  )

# View which variables have a p-value of less than 5%

hom_prop_df %>% 
  dplyr::filter(p_value < 0.05) %>% 
  view()

```

# plot of different probabilities of using a drug by class

```{r}

no_outcome_df %>% 
  dplyr::mutate(
    class = lc_list[[4]]$predclass
  ) %>% 
  dplyr::select(alcflg, herflg, marflg, mthamflg, class) %>% 
  dplyr::group_by(class) %>% 
  dplyr::mutate(
    total_class = n()
  ) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_longer(
    cols = c(alcflg, herflg, marflg, mthamflg), 
    names_to = "variable", 
    values_to = "response"
  ) %>% 
  dplyr::group_by(variable, response, class) %>% 
  dplyr::mutate(
    class_response = n()
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::distinct() %>% 
  dplyr::filter(response == "Yes") %>% 
  dplyr::mutate(
    freq = class_response/total_class
  ) %>% 
  dplyr::select(variable, class, freq) %>% 
  dplyr::bind_rows(
    tibble(
      variable = "alcflg", 
      class = 3, 
      freq = 0
    )
  ) %>%
  dplyr::arrange(variable, class) %>% 
  dplyr::mutate(
    class = factor(class)
  ) %>% 
  ggplot2::ggplot(aes(x = variable, y = freq, group = class, color = class)) +
  ggplot2::geom_point() +
  ggplot2::geom_path()

```



